---
title: "Exploratory Data Analysis in R - Statistics in R Datacamp"
author: "Carter Wolff"
date: "2023-06-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---Exploring Categorical Data
More Cleaning and Initial Review of Data
-One practice of cleaning and exploring data is reviewing the levels of a variable. The levels are the different outcomes of a variable. For example, size can have multiple categorical levels like small, medium, and large. The levels() allows you to view all the levels for a given variable.
-dplyr contains functions to manipulate these levels, such as the droplevels() function. droplevels() only filters out levels that are not in a data set, meaning the levels you wish to drop should be filtered out using filter() before using the droplevels() function.

Counts versus Proportions
-prop.table() can convert the table() results into proportions. In the standard prop.table() call, all values across all rows sum to 1. 
-conditional proportions evaluate data strictly across rows or columns. This means that the sum of a row or column is 1 rather than the entire table.
  -To specify rows, include a 1 after the table argument in prop.table(); for columns, include a 2:
    prop.table(tablename, 1) is condition on rows; prop.table(tablename, 2) is conditions on columns
  -conditional proportions are good to evaluate the association between two variables.

---Exploring Numerical Data
Visualizing data in higher dimensions
-facet_wrap or facet_grid gives options to plot by a third variable, as does using the multiple shape/color/fills options within ggplot

---Email Case study
-Using data from the openinto package in R

```{r}
# Load packages
library(dplyr)
library(ggplot2)
library(openintro)

# Compute summary statistics for spam and not spam emails
email %>%
  group_by(spam) %>%
  summarise(median(num_char), IQR(num_char))

# Create plot to compare number of characters in spam and non spam emails where number of characters is log transformed
email %>%
  mutate(log_num_char = log(num_char)) %>%
  ggplot(aes(x = spam, y = log_num_char)) +
  geom_boxplot()

###comparison of exclamation marks used in spam and non spam emails. 
# Compute center and spread for exclaim_mess by spam
email %>%
  group_by(spam) %>%
  summarize(median(exclaim_mess),
            IQR(exclaim_mess))

# Create plot for spam and exclaim_mess
email %>%
  mutate(log_exclaim_mess = log(exclaim_mess + 0.01)) %>%
  ggplot(aes(x = log_exclaim_mess)) +
  geom_histogram() +
  facet_wrap(~ spam)

# Alternative plot: side-by-side box plots
email %>%
  mutate(log_exclaim_mess = log(exclaim_mess + 0.01)) %>%
  ggplot(aes(x = 1, y = log_exclaim_mess)) +
  geom_boxplot() +
  facet_wrap(~ spam)

# Alternative plot: Overlaid density plots
email %>%
  mutate(log_exclaim_mess = log(exclaim_mess + .01)) %>%
  ggplot(aes(x = log_exclaim_mess, fill = spam)) +
  geom_density(alpha = 0.3)
```

Approach to zero inflated datasets
-morph into categorical variable with two levels. Zero and non-zero. Can aproach the analyses of these separately. 