---
title: "Statistics in R - Data Camp"
author: "Carter Wolff"
date: "2023-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

If Statements: 
```{r}
number <- 100

if (number < 10) {
  if (number < 5) {
    result <- "extra small"
  } else {
    result <- "small"
  }
} else if (number < 100) {
  result <- "medium"
} else {
  result <- "large"
}
print(result)
```

Writing Functions: 
   "Function name" <- function(arg1, arg2){
   body of function
}
-Setting argument equal to something (i.e., arg2 = FALSE) will specify a default value
     "Function name" <- function(arg1, arg2 = FALSE){
   body of function
}
EXAMPLES of Writing Functions
```{r}
pow_two <- function(x) {
  y <- x ^ 2
  print(paste(x, "to the power two equals", y))
  return(y)
}

#incorporating a default assignment to argument 2 ("print_info")
# Finish the pow_two() function
pow_two <- function(x, print_info = TRUE) {
  y <- x ^ 2
  if (print_info == TRUE){
      print(paste(x, "to the power two equals", y))
  }
  return(y)
}

#r cannot manipulate variables INSIDE a function
triple <- function(x) {
  x <- 3*x
  x
}
a <- 5
triple(a)
a

###examples of functions with vectors
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Define the interpret function
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

interpret(linkedin[1])
interpret(facebook[2])

###nested functions with vectors

# The interpret() can be used inside interpret_all()
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}
# Define the interpret_all() function
# views: vector with data to interpret
# return_sum: return total number of views on popular days?
interpret_all <- function(views, return_sum = TRUE) {
  count <- 0

  for (v in views) {
    count <- count + interpret(v)
  }
  if (return_sum == TRUE) {
    return(count)
  } else {
    return(NULL)
  }
}
# Call the interpret_all() function on both linkedin and facebook
interpret_all(linkedin)
interpret_all(facebook)

```

R Packages Section
-installing packages: 
  -install.packages("package name")
  
-load package (attaches it to search list for r to use while executing code)
  -library("package name")
  -require() - also loads packages, but will install packages that are not installed (something that library() does not do)
    -Will return FALSE if desired package fails installation

lapply in R
-function to apply a function to each list element within a list
-elements of each list are maintained
-lapply can also be used with anonymous functions (functions where the name is not specified )
```{r}
###applies lowercase letters to the names in pioneers
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
#split elements in pioneers vector by ":"
split <- strsplit(pioneers, split = ":")
#uses the "tolower" function to apply lowercase to each element in the list "split"
split_low <- lapply(split, tolower)


###using lapply for user created functions to select first and second value for each list element
# Code from previous exercise:
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# function to select the first value of each list element
select_first <- function(x) {
  x[1]
}

names <- lapply(split_low, select_first)

# function to select second value to each list element
select_second <- function(x){
  x[2]
}

years <- lapply(split_low, select_second)

# replaces "select_first" function inside lapply

names <- lapply(split_low, function(x) {x[1]})

# replaces "select_second" function inside lapply
years <- lapply(split_low, function(x) {x[2]})
```

sapply (simplified apply) function in R
-lapply works well with lists because lists can contain multiple types of R objects (logical, character, numeric, etc)
-if all elements are of the same object type, you can use sapply to convert the result of the function to an array
-sapply uses lapply behind the scene, then unlist the result of the function. Similar to "unlist(lapply(x, fun()))"
-sapply cannot simplify elements that are of different lengths. Thus, R returns a value similar to that of lapply

```{r}
#creating temp list that is used in the DataCamp Exercise
temp <- list(c(3, 7, 9, 6, -1), c(6, 9, 12, 13, 5), c(4, 8, 3, -1, -3), c(1, 4, 7, 2, -2), c(5, 7, 9, 4, 2), c(-3, 5, 8, 9, 4), c(3, 6, 9, 4, 1))

# Use lapply() to find each day's minimum temperature
lapply(temp, min)

# Use sapply() to find each day's minimum temperature
sapply(temp, min) #returns the same value as unlist(lapply(temp, min))

####sapply with user created funtions
# Create a function that returns min and max of a vector: extremes
extremes <- function(x) {
  c(min = min(x), max = max(x))
}

# Apply extremes() over temp with sapply()
sapply(temp, extremes) #extremes returns two values per element in temp (min and max). Thus, a 2D array is created instead of a 1D array

# Apply extremes() over temp with lapply()
lapply(temp, extremes)

### What happens when sapply cannot simplify

# Function to return all values below 0 from temp
below_zero <- function(x) {
  return(x[x < 0])
}

# Apply below_zero over temp using sapply(): freezing_s
freezing_s <- sapply(temp, below_zero)

# Apply below_zero over temp using lapply(): freezing_l
freezing_l <- lapply(temp, below_zero)

# Are freezing_s and freezing_l identical?
identical(freezing_s, freezing_l) #since sapply cannot simply list elements of different lengths, freezing_s and freezing_l return the exact same values, resulting in identical() being TRUE
```

vapply in R
-sapply attempst to apply a list to an array, but sometimes it is not possible to simplify. Thus you may return a result similar to lapply, which can be dangerous
-vapply requires you to specify a specific output format (defined as FUN.VALUE i.e., FUN.VALUE = numeric(1) or FUN.VALUE = charter(2))
-vapply is usually the safer choice, as it requires you to know the output format of your function beforehand
```{r}
#using temp again for this exercise
temp <- list(c(3, 7, 9, 6, -1), c(6, 9, 12, 13, 5), c(4, 8, 3, -1, -3), c(1, 4, 7, 2, -2), c(5, 7, 9, 4, 2), c(-3, 5, 8, 9, 4), c(3, 6, 9, 4, 1))

#create a function to return the min, avg, and max temperature of each list element
basics <- function(x) {
  c(min = min(x), mean = mean(x), max = max(x))
}

# Apply basics() over temp using vapply()
vapply(temp, basics, numeric(3)) #here we specify that the output of the function should be a numerical value of size 3 (for min, avg, and max)

# Don't edit these two lines
vec1 <- c(1.5, 2.5, 8.4, 3.7, 6.3)
vec2 <- rev(vec1)

mean(abs(vec1), abs(vec2), trim = 0)
```
Date and Time Formats in R

%Y: 4-digit year (1982)
%y: 2-digit year (82)
%m: 2-digit month (01)
%d: 2-digit day of the month (13)
%A: weekday (Wednesday)
%a: abbreviated weekday (Wed)
%B: month (January)
%b: abbreviated month (Jan)
